### 2007-01-09 Documentation/part2.tex
*** empty log message ***

The description of the neutrino oscillation physics is, at least numerically, relatively simple. We use the {\em evolution operator method}~\cite{Ohlsson:1999um}  to compute the neutrino oscillation probabilities and divide the matter density profile into layers of constant matter density. For each of these layers, the Hamiltonian in matter is diagonalized in order to propagate the neutrino transition amplitudes. Finally, the transition probability is obtained by the absolute square of the total neutrino transition amplitudes. Depending on the precision of the studied experiment, this approach turns out to be precise enough in Earth matter even if only a small number matter density steps is used. Since we allow an uncertainty of the matter density profile, it is, in fact, in most cases sufficient to consider only one density step with the average matter density together with a matter density uncertainty~\cite{Ohlsson:2003ip}. Note that this approach may not be applicable to quickly varying extraterrestrial matter density profiles.

@@

The first line of every experiment definition file has to be
{\tt !\%GLoBES}
in order not to confuse it with some other file format.

@@

{\tt \tb @signalerror = 0.001 : 0.0001}\\

@@

{\tt \tb @backgroundcenter = 1 : 0.0}\\
{\tt \tb @backgrounderror = 0.05 : 0.0001}\\
{\tt \tb @errordim\_sys\_on = 0}\\
{\tt \tb @errordim\_sys\_off = 2}\\

@@

efficiency of $0.45$. The signal error variable has two components: 
The first one is the normalization error of the signal, here $0.1\%$. The second 

@@

use the fraction $1\cdot 10^{-5}$ of the channel named {\tt \#disappearance}, \ie , the muon neutrinos with a mis-identified charge. The background center variable allows to rescale the total background contribution from all background components
simultaneously. It is only useful if there is more than one background component, otherwise it is usually $1$. The background error variable is defined such as the signal error variable, \ie\ we have a $5\%$ background uncertainty and a very small energy calibration error. The ``error dimension variable'' {\tt @errordim\_sys\_X} selects how the systematical errors are treated (\cf, \Tab~\ref{tab:error_dim}).

@@

We now give a short introduction to the syntax of AEDL.
 The first eight characters have to be {\tt \%!GLoBES}
in order to avoid parsing megabytes of chunk
 and producing thousands of error messages. 
%
Comments can be used such as in C:
\begin{quote}
{\tt /* This starts a comment\\
 and here the comment ends */
}
\end{quote}
There are pre-defined variables which all start with {\tt \$}. Their range
is also checked. For example,  {\tt 
\$bins} can be only between $0$ and $500$.\footnote{The upper limit is 
only there for safety reasons, the memory is allocated dynamically.} If one uses a {\tt float} quantity where  an {\tt int} is expected, the {\tt float} will be converted to an {\tt int} in the same way as in C.  For example, we have scalar variables
\begin{quote}
{\tt
\$bins = 10\\
\$baseline = 1200.0
}
\end{quote}
and simple lists
\begin{quote}
{\tt
\$densitytab=\{1.0,2.2343,3.3432\} 
}
\end{quote}
%
Since there are often groups of data which we want to refer to later,
environments can be used. This is illustrated 
with the channel definition part:
\begin{quote}
{\tt channel(\#ch1)<\\
\tb  $\ldots$\\
>
}
\end{quote}
The first part is the type of environment, which is {\tt channel} here. 
There are the following types of environment in AEDL:
\begin{quote}
{\tt flux\\
cross\\
channel\\
energy\\
rule
}
\end{quote}
Besides the environment type, there is a user-defined name 
beginning with {\tt \#}
in the above example: {\tt \#ch1}. It can be used later to refer to the 
channel defined in {\tt <$\ldots$>}. Those names are so-called 
``automatic variables'' and have to start with {\tt \#}. Note that these names have to be unique and can only be refered to after their definition.
However, similar to C, one can give a declaration without definition before:
\begin{quote}
{\tt    channel(\#ch2)<>}
\end{quote}
Now one can refer to the name {\tt \#ch2}, while the actual channel definition comes later. The internal representation of this automatic
variable is a number, which obtains its value from a counter for each type of environment. For example, for {\tt channel} the counter is {\tt numofchannels}. The counter keeps track of how many different names 
there are for one type of environment, which means that it counts the number of channels, rules, energy resolution functions \etc . Thus, the automatic
variables are numbered in the order of their definition, and the number
can later be used to refer to them in the C code (from $0$ to {\tt numof...}$-1$). In order to facilitate the the mapping from names in AEDL to indices
in C there are two functions \GLB{NameToValue} and \GLB{ValueToName} which
make this transition (see \Sec~\ref{sec:aedl_names}, 
page~\pageref{sec:aedl_names}).

@@

\index{aedl}{include@{\tt include}}
As another feature of AEDL one can use include files with the {\tt include} command. Includes can be nested up to {\tt MAX\_INCLUSION\_DEPTH}, which is currently set to $10$. Error reporting works 
 for nested includes, too. The included file is not required to begin 
 with {\tt \%!GLoBES} to facilitate cut and paste:
\begin{quote}
{\tt include "./file\_1"}
\end{quote}
With this include mechanism, one can use constructions such as 
\begin{quote}
{\tt    include  "NuFact.gls"\\
        \#NEXT\#\\
        include   "JHFHK.gls"
}
\end{quote}
in order to initialize a combined analysis of the experiments defined in the files {\tt NuFact.gls} and {\tt JHFHK.gls}. Note that one has 
to use quotation marks for filenames in \AEDL.\index{norm}{File names}
Even if one uses the
automatic variable {\tt \#CC} in both experiments, 
but the cross section data are different (for example, because of different target nuclei), the correct 
cross section data will be applied to each of the experiments. 
Note that, alternatively, one can 
also load both files successively by two separate calls of 
\GLB{InitExperiment}.

@@

In addition, some simple algebraic manipulations are possible, such as
\begin{quote}
{\tt
Pi+1\\
\verb+Pi^2+\\
sin(Pi/2)\\
}
\end{quote}
The following mathematical functions from {\tt <math.h>} are available: 
{\tt sin}, {\tt cos}, {\tt tan}, {\tt asin}, {\tt acos}, {\tt atan}, 
{\tt log}, {\tt log10}, {\tt exp}, {\tt sqrt}.
\index{aedl}{sin@{\tt sin}}
\index{aedl}{cos@{\tt cos}}
\index{aedl}{tan@{\tt tan}}
\index{aedl}{asin@{\tt asin}}
\index{aedl}{acos@{\tt acos}}
\index{aedl}{atan@{\tt atan}}
\index{aedl}{log@{\tt log}}
\index{aedl}{log10@{\tt log10}}
\index{aedl}{exp@{\tt exp}}
\index{aedl}{sqrt@{\tt sqrt}}

These functions can be used everywhere, where
otherwise only a scalar number would appear. However, they can not be
applied to lists, such as {\tt sin(\{1,2,3\})} will not work.

@@

>>>>

The description of the neutrino oscillation physics is, at least numerically, relatively simple. We use the {\em evolution operator method} (see, \eg, \Ref~\cite{Ohlsson:1999um})  to compute the neutrino oscillation probabilities and divide the matter density profile into layers of constant matter density. For each of these layers, the Hamiltonian in matter is diagonalized in order to propagate the neutrino transition amplitudes. Finally, the transition probability is obtained by the absolute square of the total neutrino transition amplitudes. Depending on the precision of the studied experiment, this approach turns out to be precise enough in Earth matter even if only a small number matter density steps is used. Since we allow an uncertainty of the matter density profile, it is, in fact, in most cases sufficient to consider only one density step with the average matter density together with a matter density uncertainty~\cite{Ohlsson:2003ip}. Note that this approach may not be applicable to quickly varying extraterrestrial matter density profiles.

@@

The first line of every experiment definition file has to be
\begin{quote}
{\tt !\%GLoBES}
\end{quote}
in order not to confuse it with some other file format.
In addition, \GLOBES\ 3.0 and higher requires the identification of the
minimum \GLOBES\ version the \AEDL\ file can be used with:
\begin{quote}
{\tt \$version="3.0.0"}
\end{quote}

@@

{\tt \tb @signalerror = 0.025 : 0.0001}\\

@@

%{\tt \tb @backgroundcenter = 1 : 0.0}\\
{\tt \tb @backgrounderror = 0.2 : 0.0001}\\
{\tt \tb @sys\_on\_function = "chiSpectrumTilt"}\\
{\tt \tb @sys\_off\_function = "chiNoSysSpectrum"}\\

@@

efficiency of $0.45$ (in a more realistic simulation, one would introduce an
energy threshold function). The signal error variable has two components: 
The first one is the normalization error of the signal, here $2.5\%$. The second 

@@

use the fraction $1\cdot 10^{-5}$ of the channel named {\tt \#disappearance}, \ie , the muon neutrinos with a mis-identified charge. 
%The background center variable allows to rescale the total background contribution from all background components
%simultaneously. It is only useful if there is more than one background component, otherwise it is usually $1$. 
The background error variable is defined such as the signal error variable, \ie , we have a $20\%$ background uncertainty and a very small energy calibration error. In addition, the systematics treatment is specified
in {\tt @sys\_on\_function} (for systematics switched on) and {\tt @sys\_off\_function} (for systematics switched off) -- see \Tab~\ref{tab:error_dim}.

@@

We now give a short introduction to the syntax of AEDL.
 The first eight characters have to be {\tt \%!GLoBES}
in order to avoid parsing megabytes of chunk
 and producing thousands of error messages.  In addition, the version of the minimum
\GLOBES\ version the \AEDL\ file is running with has to be defined by {\tt \$version}, such as
{\tt \$version="3.0.0"}.
\index{aedl}{version@{\tt \$version}}
%
Comments can be used such as in C:
\begin{quote}
{\tt /* This starts a comment\\
 and here the comment ends */ \\
// Another comment
}
\end{quote}
There are pre-defined variables which all start with {\tt \$}. Their range
is also checked. For example,  {\tt 
\$bins} can be only between $0$ and $500$.\footnote{The upper limit is 
only there for safety reasons, the memory is allocated dynamically.} If one uses a {\tt float} quantity where  an {\tt int} is expected, the {\tt float} will be converted to an {\tt int} in the same way as in C.  For example, we have scalar variables
\begin{quote}
{\tt
\$bins = 10\\
\$baseline = 1200.0
}
\end{quote}
and simple lists
\begin{quote}
{\tt
\$densitytab=\{1.0,2.2343,3.3432\} 
}
\end{quote}
%
Since there are often groups of data which we want to refer to later,
environments can be used. This is illustrated 
with the channel definition part:
\begin{quote}
{\tt channel(\#ch1)<\\
\tb  $\ldots$\\
>
}
\end{quote}
The first part is the type of environment, which is {\tt channel} here. 
There are the following types of environments in AEDL:
\begin{quote}
{\tt flux\\
cross\\
channel\\
energy\\
rule
}
\end{quote}
Besides the environment type, there is a user-defined name 
beginning with {\tt \#}
in the above example: {\tt \#ch1}. It can be used later to refer to the 
channel defined in {\tt <$\ldots$>}. Those names are so-called 
``automatic variables'' and have to start with {\tt \#}. Note that these names have to be unique and can only be refered to after their definition.
However, similar to C, one can give a declaration without definition before:
\begin{quote}
{\tt    channel(\#ch2)<>}
\end{quote}
Now one can refer to the name {\tt \#ch2}, while the actual channel definition comes later. The internal representation of this automatic
variable is a number, which obtains its value from a counter for each type of environment. For example, for {\tt channel} the counter is {\tt numofchannels}. The counter keeps track of how many different names 
there are for one type of environment, which means that it counts the number of channels, rules, energy resolution functions \etc . Thus, the automatic
variables are numbered in the order of their definition, and the number
can later be used to refer to them in the C code (from $0$ to {\tt numof...}$-1$). In order to facilitate the the mapping from names in AEDL to indices
in C there are two functions \GLB{NameToValue} and \GLB{ValueToName} which
make this transition (see \Sec~\ref{sec:aedl_names}, 
page~\pageref{sec:aedl_names}).

@@

\index{aedl}{include@{\tt include}}
As another feature of AEDL one can use include files with the {\tt include} command. Includes can be nested up to {\tt MAX\_INCLUSION\_DEPTH}, which is currently set to $10$. Error reporting works 
 for nested includes, too. The included file is not required to begin 
 with {\tt \%!GLoBES} to facilitate cut and paste:
\begin{quote}
{\tt include "./file\_1"}
\end{quote}
With this include mechanism, one can use constructions such as 
\begin{quote}
{\tt    include  "Exp1.glb"\\
        \#NEXT\#\\
        include   "Exp2.glb"
}
\end{quote}
in order to initialize a combined analysis of the experiments defined in the files {\tt Exp1.glb} and {\tt Exp2.glb}. Note that one has 
to use quotation marks for filenames in \AEDL.\index{norm}{File names}
Even if one uses the
automatic variable {\tt \#CC} in both experiments, 
but the cross section data are different (for example, because of different target nuclei), the correct 
cross section data will be applied to each of the experiments. 
Note that, alternatively, one can 
also load both files successively by two separate calls of 
\GLB{InitExperiment}.

@@

In addition, some simple algebraic manipulations are possible, such as
\begin{quote}
{\tt
Pi+1\\
\verb+Pi^2+\\
sin(Pi/2)
}
\end{quote}
The following mathematical functions from {\tt <math.h>} are available: 
{\tt sin}, {\tt cos}, {\tt tan}, {\tt asin}, {\tt acos}, {\tt atan}, 
{\tt log}, {\tt log10}, {\tt exp}, {\tt sqrt}.
\index{aedl}{sin@{\tt sin}}
\index{aedl}{cos@{\tt cos}}
\index{aedl}{tan@{\tt tan}}
\index{aedl}{asin@{\tt asin}}
\index{aedl}{acos@{\tt acos}}
\index{aedl}{atan@{\tt atan}}
\index{aedl}{log@{\tt log}}
\index{aedl}{log10@{\tt log10}}
\index{aedl}{exp@{\tt exp}}
\index{aedl}{sqrt@{\tt sqrt}}
% 
These functions can be used everywhere, where
otherwise only a scalar number would appear. However, they can not be
applied to lists, such as {\tt sin(\{1,2,3\})} will not work.

@@

\section{More advanced AEDL features}

In \GLOBES\ 3.0 and higher, a number of new features can be used. The
features in this section are slighlty more advanced and can be jumped over at a first reading of
the manual. The
most important one are lists as variables in \AEDL . They start with {\tt \%}, such as
\begin{quote}
{\tt
\%effs = \{ 0.2, 0.4, 0.6, 0.8, 1.0, 1.0, 1.0 \}
}
\end{quote}
Functions can be threaded over lists, \ie, by applying the function to each element
and returning a list. Note that the original list will be destroyed by this
process. Therefore, it is necessary to create a copy of your list if you want to
use the original and the threading result. For that purpose, the {\tt copy}
funcion is provided:
\begin{quote}
{\tt
 listb = copy(lista) \\
 listb := lista // Alternative method 
}
\end{quote}
\index{aedl}{copy@{\tt copy}}
You will also need to use {\tt copy} when you assign a list to an experiment structure (see below).

Two helper functions {\tt bincenter()} 
\index{aedl}{bincenter@{\tt bincenter}} and {\tt simbincenter} \index{aedl}{simbincenter@{\tt simbincenter}} return lists with the central energies of the bins or sampling points, respectively. For example,
\begin{quote}
{\tt
 \%bc=bincenter() 
}
\end{quote}
A very useful new features is an interpolation function which can directly interpolate a number of
points and evaluate them at a different set of places. For example,
\begin{quote}
{\tt
 \%energ = \{ 4.0,20.0,50.0 \} \\
 \%effs = \{ 0.0,1.0,1.0 \} \\
 \%ires = interpolation(\%energ,\%effs,1,\%bc) 
}
\end{quote}
\index{aedl}{interpolation@{\tt interpolation}}
interpolates the points with x-values {\tt \%energ} and y-values {\tt \%effs} with the interpolation order one (linear interpolation, third parameter) and evaluates the interpolation result at the bin centers obtained above, \ie, it returns a list of the y-values at the places specified by the last parameter. The only allowed interpolation orders are ``1'' (linear) and ``2'' (cubic splines). This example creates
a neutrino factory energy threshold function linearly climbing from 0 to 1 between 4~GeV and 20~GeV. It can be directly used in a channel definition, \eg, 
\begin{quote}
{\tt
 channel(\#nu\_mu\_appearance)< \\
\hspace*{0.5cm}	@channel = \#mu\_plus: +: electron: muon: \#CC: \#MINOS\ \\
\hspace*{0.5cm}	@post\_smearing\_efficiencies = copy(\%ires) \\
>
}
\end{quote}

For debugging purposes of lists and numbers, one can \GLOBES\ now supports output directly
from \AEDL\ files:
\begin{quote}
{\tt 
 R = 1.15 \\
 echo(R) // Print without line feed \\
 line(2) // Two line feeds \\
 echon(ires) // Print with line feed 
}
\end{quote}
\index{aedl}{echo@{\tt echo}}
\index{aedl}{echon@{\tt echon}}
\index{aedl}{line@{\tt line}}
